---
title: "OralHealth"
format: html
editor: visual  
---

## Data preparation

```{r}
snp_euro <- final_clsa_european %>% filter(nchar(major_allele) == 1 & nchar(minor_allele) ==1)
chr <- snp_euro$chromosome
bp <- snp_euro$position
tas2r <- snp_euro$tas2r

write.table(data.frame(chr, bp,bp,tas2r), "snp_euro.txt", quote = F, row.names = F, col.names = F)
```

tas2r snp file for plink (n=124)

```{r}

new_data <- data.frame(snp_euro$chromosome, snp_euro$position, snp_euro$position, snp_euro$tas2r)

write.table(new_data, "snp_euro_all.txt", quote = F, row.names = F, col.names = F)
```

Load baseline data

```{r}
bl_com <- read.csv("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_CoPv7.csv") 
bl_trm <- read.csv("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_Trav4.csv") 
bl_com_dict <- read_excel("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_CoPv7-dictionary.xlsx")
bl_trm_dict <- read_excel("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_Trav4-dictionary.xlsx")
```

Extract oral health variables

```{r}
oral_variables <- bl_com_dict %>% filter(str_detect(label,"Oral health")|str_detect(label, "teeth")|str_detect(label,"eating in")|str_detect(label, "tooth"))%>% select(-table)

#include sex, age, gwas_id, and oral health variables 
oral_health <- bl_com %>%rename(gwas_id= ADM_GWAS3_COM, sex = SEX_ASK_COM, age =AGE_NMBR_COM) %>% select(gwas_id, sex, age, any_of(oral_variables$name))
#write.csv(oral_health, "oral_health.csv")
#write.csv(oral_variables  %>% clean_names(), "oral_variables.csv")
```

## Create phenotype file for plink

```{r}
#Manually changed the names to lables
oral_health <- read.csv("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/oral_health.csv") %>% clean_names()
#european_id <- european_id %>%   rename(FID=fid, IID=iid) 


#add case-defined variables
oral_health <- oral_health %>%  mutate(periodontal_diseases=case_when(bleeding_gums_around_natural_teeth==1|loose_natural_tooth==1 ~ 1, TRUE ~0)) %>% 
  rename(FID=fid, IID=iid) 
#european_id <- qc_sample_euro %>%  select(FID, IID) 


#include european ancestry
oral_health_european <- left_join(european_id, oral_health, by=c("FID", "IID")) %>% select(-x, -sex, -age) 



#modify scale, affected=2, unaffected=1, except for has_one_or_more_of_original_tooth
oral_health_binary_euro <- oral_health_european %>%  select(-general_oral_health, -has_one_or_more_of_original_teeth, -uncomfortable_eating_in_last_12_months, -avoided_eating_in_last_12_months, -frequency_brushing_teeth_and_or_dentures, -frequency_brushing_teeth_and_or_dentures_unit_of_measure, -ohp_other_specify_01, -ohp_other_specify_02, -ohp_other_specify_03, -ohp_other_specify_04, -ohp_other_specify_05) %>% mutate_all(.funs = list(~ifelse(. == 1, 2, ifelse(. == 0, 1, .)))) 

#add dental problems case definition
oral_health_binary_euro <- oral_health_binary_euro %>%  mutate(dental_problems=case_when(decayed_natural_tooth==2|toothache==2|broken_natural_tooth==2 ~ 2, decayed_natural_tooth==1|toothache==1|broken_natural_tooth==1 ~ 1, TRUE ~ NA)) %>% #temporomandibular pain
  mutate(temporomandibular_pain=case_when(sore_jaw_muscles==2| painful_jaw_joints==2 ~ 2, sore_jaw_muscles==1|painful_jaw_joints==1 ~ 1, TRUE ~ NA))

#oral_health_status and avoided/uncomfrotable eating binary case definition
oral_health_binary_euro <- left_join(oral_health_binary_euro, 
                                     oral_health_european %>% select(FID, general_oral_health, has_one_or_more_of_original_teeth, avoided_eating_in_last_12_months, uncomfortable_eating_in_last_12_months) %>% mutate(poor_general_oral_health=case_when(general_oral_health==1|general_oral_health==2|general_oral_health==3 ~ 1, TRUE ~ 2)), by="FID") %>% rename(edentulism="has_one_or_more_of_original_teeth") %>% select(-wears_dentures) %>% 
  mutate(avoided_eating=case_when(avoided_eating_in_last_12_months==1| avoided_eating_in_last_12_months==2 ~ 2, TRUE ~1)) %>% 
  mutate(uncomfortable_eating=case_when(uncomfortable_eating_in_last_12_months==1| uncomfortable_eating_in_last_12_months==2 ~ 2, TRUE ~1)) 

oral_health_binary_euro$edentulism[oral_health_binary_euro$edentulism == 8] <- NA

#avoided_eating_binary case definition

 
#Remove individuals with edentulism + reported at least one problem related to tooth
oral_health_binary_euro <- oral_health_binary_euro %>%
  filter(!(edentulism == 2 & (
    toothache == 2 |
    decayed_natural_tooth == 2 |
    loose_natural_tooth == 2 |
    broken_natural_tooth == 2 |
    sore_gums_around_natural_teeth == 2 |
    bleeding_gums_around_natural_teeth == 2
  )))%>% select(-avoided_eating_in_last_12_months, -general_oral_health, -uncomfortable_eating_in_last_12_months)

oral_health_european <- oral_health_european %>%
  filter(!(has_one_or_more_of_original_teeth == 2 & (
   toothache == 1 |
    decayed_natural_tooth == 1 |
    loose_natural_tooth == 1 |
    broken_natural_tooth == 1 |
    sore_gums_around_natural_teeth == 1 |
    bleeding_gums_around_natural_teeth == 1
  ))) %>% select(-avoided_eating_in_last_12_months, -general_oral_health, -uncomfortable_eating_in_last_12_months)

```

Dental variables

```{r}
#set variables as NA if the patient is toothless
tooth_variables <- c("toothache",
  "decayed_natural_tooth",
  "loose_natural_tooth",
  "broken_natural_tooth",
  "sore_gums_around_natural_teeth",
  "bleeding_gums_around_natural_teeth",
  "periodontal_diseases",
  "dental_problems")
oral_health_binary_euro <- oral_health_binary_euro %>%
  mutate(across(all_of(tooth_variables), 
                ~ ifelse(edentulism == 2, NA, .)))

write.table(oral_health_binary_euro,"/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/oral_health_binary_euro.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")

```

Create phenotype for for noneuropean for plink not changed for the variable names

```{r}

#include noneuropean ancestry
#noneuropean_id <- noneuropean_id %>% rename(FID=fid, IID=iid)
oral_health_noneuropean <- left_join(noneuropean_id %>% rename(FID="fid"), oral_health, by=c("FID")) %>% select(-x, -sex, -age, -iid)

#modify scale, affected=2, unaffected=1, except for has_one_or_more_of_original_tooth
oral_health_binary_noneuro <- oral_health_noneuropean %>%  select(-general_oral_health, -has_one_or_more_of_original_teeth, -uncomfortable_eating_in_last_12_months, -avoided_eating_in_last_12_months, -frequency_brushing_teeth_and_or_dentures, -frequency_brushing_teeth_and_or_dentures_unit_of_measure, -ohp_other_specify_01, -ohp_other_specify_02, -ohp_other_specify_03, -ohp_other_specify_04, -ohp_other_specify_05) %>% mutate_all(.funs = list(~ifelse(. == 1, 2, ifelse(. == 0, 1, .)))) 


#oral_health_status and avoided/uncomfrotable eating binary case definition
oral_health_binary_noneuro <- left_join(oral_health_binary_noneuro, 
                                     oral_health_noneuropean %>% select(FID, general_oral_health, has_one_or_more_of_original_teeth, avoided_eating_in_last_12_months, uncomfortable_eating_in_last_12_months) %>% mutate(poor_general_oral_health=case_when(general_oral_health==1|general_oral_health==2|general_oral_health==3 ~ 1, TRUE ~ 2)), by="FID") %>% rename(edentulism="has_one_or_more_of_original_teeth") %>% select(-wears_dentures) %>% 
  mutate(avoided_eating=case_when(avoided_eating_in_last_12_months==1| avoided_eating_in_last_12_months==2 ~ 2, TRUE ~1)) %>% 
  mutate(uncomfortable_eating=case_when(uncomfortable_eating_in_last_12_months==1| uncomfortable_eating_in_last_12_months==2 ~ 2, TRUE ~1)) 

oral_health_binary_noneuro$edentulism[oral_health_binary_noneuro$edentulism == 8] <- NA


#Remove individuals with edentulism + reported at least one problem related to tooth
#oral_health_binary_noneuro <- oral_health_binary_noneuro %>%
 # filter(!(edentulism == 2 & (
  #  toothache == 2 |
   #decayed_natural_tooth == 2 |
    #loose_natural_tooth == 2 |
    #broken_natural_tooth == 2 |
    #sore_gums_around_natural_teeth == 2 |
    #bleeding_gums_around_natural_teeth == 2
#  )))


write.table(oral_health_binary_noneuro %>% select(FID, IID, sore_jaw_muscles),"/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/sore_jaw_muscles_noneuro.txt", quote = FALSE, row.names = FALSE, sep = "\t")

```

## Chi-squared test with plink

```{bash}

DATA=/home/mojan/Mojan_data/Genomics3_clsa/output/

#create bed files for the different phenotypes
#If phenotype file contains multiple phenotypes, use --pheno-name to specify phenotype name to be used
#--all-pheno for all phenotypes for the final analysis (--assoc)

#association test for oral health phenotypes
plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex \
          --pheno $DATA/oral_health/oral_health_binary_euro.txt \
          --assoc --ci 0.95 --adjust --all-pheno\
          --out $DATA/oral_health/univariable_results/gwas


```

## Summarize plink chi-squared outcomes

```{r}
final_clsa_european <- left_join(clsa_european, consequence_filtered_european %>% select(snp, rs_id, symbol, consequence, eur_af, p_value, protein_position, amino_acids, sift, poly_phen, strand, type, id), by="snp")


phenotypes <- c(
  "toothache",
  "cannot_chew_adequately",
  "swelling_in_mouth",
  "dry_mouth",
  "burning_mouth",
  "sore_jaw_muscles",
  "painful_jaw_joints",
  "decayed_natural_tooth",
  "loose_natural_tooth",
  "broken_natural_tooth",
  "sore_gums_around_natural_teeth",
  "bleeding_gums_around_natural_teeth",
  "denture_related_sores",
  "dirty_teeth_or_dentures",
  "bad_breath",
  "periodontal_diseases",
  "edentulism",
  "poor_general_oral_health",
  "avoided_eating",
  "uncomfortable_eating",
  "no_oral_health_problems",
  "dental_problems",
  "temporomandibular_pain"
)
```

```{r}
gwas_outcomes_oral <- list()
gwas_combined <- read.table("~/Mojan_data/Genomics3_clsa/output/oral_health/univariable_results/gwas.toothache.assoc",header = TRUE) %>% 
  clean_names %>% select(snp, p) %>% mutate_all(tolower)%>% left_join(., snp_euro_oral %>% select(snp, gene, rs_id, consequence), by="snp")

gwas_combined_significant <- read.table("~/Mojan_data/Genomics3_clsa/output/oral_health/univariable_results/gwas.toothache.assoc",header = TRUE) %>% 
  clean_names %>% select(snp, p) %>% mutate_all(tolower) %>% left_join(., snp_euro_oral, by="snp")

# L0.002088oop through each phenotype
for (phenotype in phenotypes) {
  # Read the corresponding file
  file_path <- paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/univariable_results/gwas.", phenotype, ".assoc")
  data <- read.table(file_path, header = TRUE)%>% clean_names() %>%
    mutate_all(tolower) %>% mutate(adjusted_p=as.numeric(p)*35)
  
  # Assign the dataframe to the list with the same name as phenotype
  gwas_outcomes_oral[[phenotype]] <- left_join(data, snp_euro_oral %>% select(snp, gene, rs_id, consequence), by="snp")
  
    gwas_combined_significant <- left_join(gwas_combined_significant, data %>% filter(adjusted_p<0.06)%>% select(snp, adjusted_p, or, l95, u95) %>% 
    rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, 
           !!paste0("or_", phenotype) := or,
           !!paste0("l95_", phenotype) := l95,
           !!paste0("u95_", phenotype) := u95),
    by="snp")

  
  gwas_combined <- left_join(gwas_combined, data %>% 
                               select(snp, adjusted_p, or, l95, u95) %>% 
    rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, 
           !!paste0("or_", phenotype) := or,
           !!paste0("l95_", phenotype) := l95,
           !!paste0("u95_", phenotype) := u95),
    by="snp")
  
}
gwas_combined_significant <- gwas_combined_significant[rowSums(!is.na(gwas_combined_significant %>% select(-snp, -p, -gene, -rs_id, -consequence))) > 0, ]

  
  
write.csv(gwas_combined, "gwas_combined.csv", col.names = TRUE)
write.csv(gwas_combined_significant, "gwas_combined_significant.csv", col.names = TRUE)
```

## Prevalence of outcomes

```{r}

library(gtsummary)
#used oral_health_european because also includes categorical variables of multiple categories
oral_health_european <- left_join(oral_health_european, bl_com %>% rename(IID= ADM_GWAS3_COM, sex = SEX_ASK_COM, age =AGE_NMBR_COM) %>%  select(IID, sex, age), by= "IID") %>% mutate(sex=case_when(sex=="M"~1, sex=="F"~2))


oral_health_european %>%  tbl_summary(include = everything(), by=sex, 
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary.docx") 

#for covariates
covariates_oral %>%  tbl_summary(include =  c(smoking, age, brushing, education, general_health, household_income), by=sex,
    missing_text = "(Missing)") %>%
    add_p() %>% 
 as_gt() %>%
  gt::gtsave(filename = "covariates_oral.docx") 

#gt summary for the total european(not by sex)
oral_health_binary_euro %>%  tbl_summary(include = everything(), 
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary_european.docx") 

oral_health_binary_euro %>%  tbl_summary(include = everything(), by=Sex,
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary_european_by_sex.docx") 


```

## Calculate the number of independent phenotypes

```{r}

library(psych)

phi_matrix <- tetrachoric(oral_health_binary_euro %>% select(-FID, -IID, -ohp_other, -ohp_no_answer, -ohp_refused, -temporomandibular_pain, -dental_problems) %>% rename(bleeding_gums=bleeding_gums_around_natural_teeth, sore_gums=sore_gums_around_natural_teeth))$rho
png("correlation_plot.png", width = 15, height = 15, units = "in", res = 450)
corrplot(phi_matrix, method = "color", type = "lower", 
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", 
         tl.cex = 1.7, 
         number.cex = 0.7, cl.cex = 1.8)

dev.off()


```

## Manhattan plot for chi-squared test results

```{r}
# Set the size of the graphic device
dev.new(width=11, height=4.5)

for (phenotype in phenotypes) {

# Prepare the data
gwas_results <- read.table(paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/univariable_results/gwas.", phenotype, ".assoc"), header = TRUE) %>% clean_names() %>% rename(chromosome=chr) %>%
  mutate_all(tolower) %>% mutate(p=as.numeric(p)) %>% mutate(chromosome=as.numeric(chromosome)) 

gwas_results <- inner_join(gwas_results, snp_euro, by=c("chromosome", "snp")) 

gwas_results <- gwas_results %>%
  mutate(log_p = -log10(p)) %>% 
  arrange(chromosome, bp) %>% 
  mutate(label = paste(snp, gene, consequence, sep = "_"))  # create new LABEL column

# Create an ordered factor for GENE based on the order in the arranged dataframe
gwas_results$gene <- factor(gwas_results$gene, levels = unique(gwas_results$gene)) 

# Set a color palette with a large number of colors
num_colors <- length(unique(gwas_results$gene))
my_palette <- rainbow(num_colors)

# Shuffle the palette
set.seed(123)  # for reproducible results
my_palette <- sample(my_palette)

# Create a scatter plot
manhattan_plot = ggplot(gwas_results, aes(x = gene, y = log_p, color = gene)) +
  geom_point(alpha = 0.5, size=3.5) +
  scale_color_manual(values = my_palette) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 2.8, linetype = "dashed", color = "red") +
  geom_text(data = subset(gwas_results, log_p > 3.5), aes(label = label), size = 5, vjust = 0, hjust = 0,check_overlap = F) +
  # ggrepel::geom_text_repel(data = subset(gwas_results, log_p_bonferroni > 4), aes(label = LABEL), 
  #                        nudge_x = 0., nudge_y = 0.0, size = 3, check_overlap = F) +
  labs(title = gsub("_|ohp", " ", phenotype), x = NULL, y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 30), plot.title = element_text(size = 30, face = "bold")) +
  guides(color = FALSE) +
  scale_y_continuous(limits = c(0, 3.5))
# Save the plot
ggsave(paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/chi_squared_manhattan_plots/", phenotype, ".png"), plot = manhattan_plot, width = 11, height = 4.5, dpi = 450)
}
```

compare males with females

```{r}

males <- c(859, 5553, 4659, 3414, 4407, 2399, 545, 141, 193, 717, 2135, 7857, 143, 399, 1031, 9330, 629, 1410, 799, 465, 1689, 102, 396, 404, 1798, 510, 1278, 754, 1094, 203, 386, 823, 4754, 1508, 3298, 610)
females <- c(1013, 4816, 5089, 3958, 4318, 2210, 447, 120, 296, 931, 2197, 7631, 213, 634, 1313, 8893, 686, 1306, 933, 535, 2196, 207, 788, 868, 1352, 522, 1124, 894, 1315, 262, 239, 735, 4618, 1700, 2831, 1198)

categories <- list(
  smoking_status = matrix(c(859, 1013, 5553, 4816, 4659, 5089), ncol=2),
  
  general_oral_health = matrix(c(
    3414, 3958, 4407, 4318, 2399, 2210, 545, 447, 141, 120
  ), ncol=2),

  uncomfortable_eating = matrix(c(
    193, 296, 717, 931, 2135, 2197, 7857, 7631
  ), ncol=2),
  

  avoided_eating = matrix(c(
    143, 213, 399, 634, 1031, 1313, 9330, 8893
  ), ncol=2),
  
  edentulism = matrix(c(629, 686), ncol=2),
  toothache = matrix(c(1410, 1306), ncol=2),
  cannot_chew = matrix(c(799, 933), ncol=2),
  swelling_in_mouth = matrix(c(465, 535), ncol=2),
  dry_mouth = matrix(c(1689, 2196), ncol=2),
  burning_mouth = matrix(c(102, 207), ncol=2),
  jaw_muscles_sore = matrix(c(396, 788), ncol=2),
  jaw_joints_painful = matrix(c(404, 868), ncol=2),
  natural_tooth_decayed = matrix(c(1798, 1352), ncol=2),
  natural_tooth_loose = matrix(c(510, 522), ncol=2),
  natural_tooth_broken = matrix(c(1278, 1124), ncol=2),
  gums_sore = matrix(c(754, 894), ncol=2),
  gums_bleed = matrix(c(1094, 1315), ncol=2),
  denture_sores = matrix(c(203, 262), ncol=2),
  teeth_dentures_dirty = matrix(c(386, 239), ncol=2),
  bad_breath = matrix(c(823, 735), ncol=2),
  no_oral_health_problems = matrix(c(4754, 4618), ncol=2),
  periodontal_disease = matrix(c(1508, 1700), ncol=2),
  dental_problem = matrix(c(3298, 2831), ncol=2),
  TMD = matrix(c(610, 1198), ncol=2)
)

# Perform Chi-square tests
p_values <- sapply(categories, function(x) {
  chisq.test(x)$p.value
})

# Bonferroni correction
alpha <- 0.05
bonferroni_alpha <- alpha / length(p_values)

# Adjust p-values (Bonferroni correction)
p_values_adjusted <- p.adjust(p_values, method = "bonferroni")

# Create a table with original and adjusted p-values
results_table <- data.frame(
  Characteristic = names(categories),
  `p-value` = p_values,
  `p-value (adjusted)` = p_values_adjusted
)


write.csv(results_table, "summary_compare_sex.csv")

#Age
hist(covariates_oral$Age, main = "Histogram of Age", xlab = "Age", col = "lightblue", border = "black")

# Q-Q Plot
qqnorm(covariates_oral$Age)
qqline(covariates_oral$Age, col = "red")

# Perform two-sample t-test 

age_male <- covariates_oral$Age[covariates_oral$Sex == 1]
age_female <- covariates_oral$Age[covariates_oral$Sex == 2]

covariates_oral %>% group_by(Sex) %>% summarize(
     Mean = mean(Age, na.rm = TRUE),
    SD = sd(Age, na.rm = TRUE)
  )

#Mann-Whitney U test (Wilcoxon rank-sum test)
wilcox.test(age_male, age_female)

```

## Create covariate files for plink

```{r}
#Sex 1=male; 2=female; other=unknown
covariates_oral <- bl_com %>% 
select(
    SEX_ASK_COM, 
    AGE_NMBR_COM,
    ADM_GWAS3_COM, 
    ORH_BRUSH_NB_MCQ,
    ORH_BRUSH_UNIT_MCQ,
    contains("SDC_ETHN"), 
    contains("SDC_CULT"), 
    ED_UDR04_COM, 
    WEA_INCNEEDS_MCQ,
    ICQ_SMOKE_COM,
    INC_TOT_COM,
    ED_UDR04_COM,
    GEN_HLTH_COM
    ) %>%
  rename(IID=ADM_GWAS3_COM, household_income=INC_TOT_COM, education=ED_UDR04_COM, general_health=GEN_HLTH_COM, age=AGE_NMBR_COM, sex=SEX_ASK_COM)

covariates_oral <- left_join(high_sample_id %>% rename(IID="iid"), covariates_oral, by="IID") %>% mutate(FID=IID) %>% mutate(sex=case_when(sex=="M" ~ 1, sex=="F" ~ 2, TRUE ~ 0)) %>% 
  mutate(household_income = ifelse(household_income %in% c(7, 8, 9), NA, household_income)) %>% 
  mutate(general_health=ifelse(general_health %in% c(8,9), NA, general_health)) %>% 
  mutate(education=ifelse(education==9, NA, education)) %>% 
  mutate(brushing=case_when(ORH_BRUSH_UNIT_MCQ==1 & ORH_BRUSH_NB_MCQ==1~ 1,
         ORH_BRUSH_UNIT_MCQ==1 & ORH_BRUSH_NB_MCQ!=1 ~ 2,
         ORH_BRUSH_NB_MCQ==998 | ORH_BRUSH_NB_MCQ==999 ~ NA,
         TRUE ~ 0)) %>% 
  mutate(smoking=case_when(ICQ_SMOKE_COM==1 ~ 2, #Current somker=2
                           ICQ_SMOKE_COM==2 ~ 0, #Never somoker=0
                           ICQ_SMOKE_COM==3 ~1)) #Former smoker=1

covariates_oral <- inner_join(covariates_oral, oral_health_binary_euro %>% select(IID), by="IID") %>% 
  select(FID, IID, age, smoking, brushing, sex, household_income, general_health, education) %>% 
   distinct(IID, .keep_all = TRUE)

write.table(covariates_oral, "/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/covariates_oral.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")


#covariates noneuro
covariates_oral_noneuro <- inner_join(covariates_oral, noneuropean_id %>% rename(IID="iid") %>%  select(-fid), by="IID") %>% 
  select(FID, IID, age, smoking, brushing, sex, household_income, general_health, education)

write.table(covariates_oral_noneuro, "/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/covariates_oral_noneuro.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")




#write.table(covariates_oral %>% select(FID, IID, Sex), "update_sex.txt", quote = FALSE, row.names = FALSE, sep = "\t")
```

## Assess multicollinearity

```{r}

png("correlation_confounders.png", width = 15, height = 15, units = "in", res = 450)
corrplot(cor(covariates_oral %>% select(age, sex, smoking, household_income, brushing, general_health, education ), use="complete.obs"), method="circle", # Increase space for labels
         type="lower",
         tl.col = "black", tl.srt = 45, 
         tl.cex = 2, 
         addCoef.col = "black", 
         number.cex = 1.6, cl.cex = 2)

dev.off()

```

## Logistic regression with PLINK

```{bash}
DATA=/home/mojan/Mojan_data/Genomics3_clsa/output

#logistic regression

plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex\
          --logistic --pheno $DATA/oral_health/oral_health_binary_euro.txt --all-pheno\
          --covar $DATA/oral_health/covariates_oral.txt \
          --covar-name age, smoking, brushing, sex, household_income, general_health, education\
          --ci 0.95 --out $DATA/oral_health/logistic_results/gwas
          
#plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex --recode --out $DATA/oral_health/snp_t2r_european_sm_qc_maf0.01_updated_sex

#noneuropean
plink --bfile $DATA/snp_t2r_noneuropean_sm_qc_maf0.01_updated_sex\
          --logistic --pheno $DATA/oral_health/sore_jaw_muscles_noneuro.txt --all-pheno\
          --covar $DATA/oral_health/covariates_oral_noneuro.txt \
          --covar-name age, smoking, brushing, sex, household_income, general_health, education\
          --ci 0.95\
          --out $DATA/oral_health/logistic_results_noneuro/sore-jaw-muscles

```

## Summarize logistic regression results

```{r}
snp_euro_oral <- read_excel("snp_euro_oral.xlsx", sheet="snp_euro_oral")
logistic_results_oral <- list()
logistic_combined_oral <- read.table("~/Mojan_data/Genomics3_clsa/output/oral_health/logistic_results/gwas.toothache.assoc.logistic",header = TRUE) %>% 
  clean_names %>% mutate_all(tolower) %>% filter(test=="add")  %>%  left_join(., snp_euro_oral, by="snp")

logistic_combined_oral_significant <- read.table("~/Mojan_data/Genomics3_clsa/output/oral_health/logistic_results/gwas.toothache.assoc.logistic",header = TRUE) %>% clean_names %>% mutate_all(tolower) %>% filter(test=="add")  %>%  left_join(., snp_euro_oral, by="snp") %>% select(-p, -or, -test, -stat)


# Loop through each phenotype
for (phenotype in phenotypes) {
  # Read the corresponding file
  file_path <- paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic")
  data <- read.table(file_path, header = TRUE)%>% clean_names() %>%
    mutate_all(tolower) %>% filter(test=="add") %>% mutate(adjusted_p=as.numeric(p)*35)
  
  # Assign the dataframe to the list with the same name as phenotype
  logistic_results_oral[[phenotype]] <- left_join(data, snp_euro_oral, by="snp")
  
   logistic_combined_oral_significant <- left_join(logistic_combined_oral_significant, data %>% filter(adjusted_p<0.06)%>%  select(snp, adjusted_p, or, l95, u95) %>% 
    rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, 
           !!paste0("or_", phenotype) := or,
           !!paste0("l95_", phenotype) := l95,
           !!paste0("u95_", phenotype) := u95),
    by="snp")

  
  logistic_combined_oral <- left_join(
  logistic_combined_oral, 
  data %>% 
    select(snp, adjusted_p, or, l95, u95) %>% 
    rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, 
           !!paste0("or_", phenotype) := or,
           !!paste0("l95_", phenotype) := l95,
           !!paste0("u95_", phenotype) := u95),
  by = "snp"
)

  
}

 #logistic_combined_oral_significant <-  logistic_combined_oral_significant[rowSums(!is.na(
   #logistic_combined_oral_significant%>% select(-snp, -p, -tas2r, -rs_id, -symbol, -consequence))) > 0, ]

write.csv(logistic_combined_oral, "logistic_combined_oral.csv", col.names = TRUE)
write.csv(logistic_combined_oral_significant, "logistic_combined_oral_significant.csv", col.names = TRUE)

#write.csv(inner_join(read_excel("logistic_combined_oral_significant.xlsx", sheet="t2r4_14_20"), snp_euro, by="snp"), "logistic_combined_oral_significant.csv", col.names = TRUE)
```

Summarize logistic results for noneuropeans

```{r}
logistic_noneuro <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/logistic_results_noneuro/sore-jaw-muscles.sore_jaw_muscles.assoc.logistic",header = TRUE) %>% filter(TEST=="ADD")
```

Manhattan plots

```{r}
# Set the size of the graphic device
dev.new(width=11, height=4.5)

for (phenotype in phenotypes) {

# Prepare the data
gwas_results <- read.table(paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic"), header = TRUE) %>% clean_names() %>% rename(chromosome=chr) %>%
  mutate_all(tolower) %>% mutate(p=as.numeric(p)) %>% mutate(chromosome=as.numeric(chromosome)) %>%
  filter(test=="add")

gwas_results <- inner_join(gwas_results, snp_euro_oral %>% select(-chromosome), by=c("snp")) 

gwas_results <- gwas_results %>%
  mutate(log_p = -log10(p)) %>% 
  arrange(chromosome, bp) %>% 
  mutate(label = paste(snp, gene, consequence, sep = "_"))  # create new LABEL column

# Create an ordered factor for GENE based on the order in the arranged dataframe
gwas_results$gene <- factor(gwas_results$gene, levels = unique(gwas_results$gene))

num_colors <- length(unique(gwas_results$gene))
my_palette <- rainbow(num_colors)

set.seed(123) 
my_palette <- sample(my_palette)

# Create a scatter plot
manhattan_plot = ggplot(gwas_results, aes(x = gene, y = log_p, color = gene)) +
  geom_point(alpha = 0.5, size = 3.5) +
  scale_color_manual(values = my_palette) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 2.8, linetype = "dashed", color = "red") +
  geom_text(data = subset(gwas_results, log_p > 3.5), aes(label = label), size = 6, vjust = 0, hjust = 0,check_overlap = F) +
  # ggrepel::geom_text_repel(data = subset(gwas_results, log_p_bonferroni > 4), aes(label = LABEL), 
  #                        nudge_x = 0., nudge_y = 0.0, size = 3, check_overlap = F) +
  labs(title = gsub("_|ohp", " ", phenotype), x = NULL, y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 30),  plot.title = element_text(size = 30, face = "bold")) +
  guides(color = FALSE) +
  scale_y_continuous(limits = c(0, 4))

# Save the plot
ggsave(paste0("~/Mojan_data/Genomics3_clsa/output/oral_health/logitmanhattan_plots/", phenotype, ".png"), plot = manhattan_plot, width = 11, height = 4.5, dpi = 450)
}
```

TAS2R20

```{r}
all_gwas_tas2r20 <- data.frame()

# Loop through phenotypes to gather the data
for (phenotype in phenotypes) {
  
  gwas_results <- read.table(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic"), header = TRUE) %>% 
    clean_names() %>% 
    rename(chromosome = chr) %>%
    mutate_all(tolower) %>% 
    mutate(p = as.numeric(p)) %>% 
    mutate(chromosome = as.numeric(chromosome)) %>% filter(test=="add")
  
  gwas_results <- inner_join(gwas_results, snp_euro, by = c("chromosome", "snp")) %>%
    mutate(log_p = -log10(p)) %>% 
    arrange(as.numeric(chromosome), bp) %>% 
    mutate(label = paste(snp, tas2r, consequence, sep = "_"))

  gwas_tas2r20 <- gwas_results %>% 
    filter(tas2r == "tas2r20") %>%
    mutate(phenotype = phenotype)
  
  all_gwas_tas2r20 <- bind_rows(all_gwas_tas2r20, gwas_tas2r20)
}

# Create a scatter plot
manhattan_plot <- ggplot(all_gwas_tas2r20, aes(x = phenotype, y = log_p, color = rs_id)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  labs(y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 27),  plot.title = element_text(size = 22, face = "bold")) +
  scale_y_continuous(limits = c(0, 2.5))

# Save the plot
ggsave("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/tas2r20/combined_plot.png", plot = manhattan_plot, width = 10, height = 6, dpi = 450)

```

## linkage disequilibrium heatmap

Chromosome 12

```{r}
ld_oral <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/independent_markers/r_squared_no_filter.ld", header = T) %>% 
  mutate(R2 = round(R2, 1)) %>% filter(CHR_A == 12)

snp_euro <- read_excel("snp_euro_oral.xlsx", sheet="snp_euro_oral") %>%   
  mutate(label = paste(rs_id, gene, sep = ","))

ld_oral <- ld_oral %>%
  left_join(snp_euro %>% filter(chromosome == 12) %>% select(snp, rs_id), by = c("SNP_A" = "snp")) %>%
  rename(rs_id_A = rs_id) %>%
  left_join(snp_euro %>% filter(chromosome == 12) %>% select(snp, rs_id), by = c("SNP_B" = "snp")) %>%
  rename(rs_id_B = rs_id)

# Create a vector of unique SNP labels
snp_labels <- (snp_euro %>% filter(chromosome == 12))$rs_id

# Create a matrix of R2 values, setting missing pairs as NA
n_snps <- length(snp_labels)
ld_matrix <- matrix(NA, nrow = n_snps, ncol = n_snps)  # Initialize with NA for missing data
rownames(ld_matrix) <- colnames(ld_matrix) <- snp_labels

for (i in 1:nrow(ld_oral)) {
  idx_a <- which(snp_labels == ld_oral$rs_id_A[i])
  idx_b <- which(snp_labels == ld_oral$rs_id_B[i])
  ld_matrix[idx_a, idx_b] <- ld_matrix[idx_b, idx_a] <- ld_oral$R2[i]
}

# Define the list of rs_ids to highlight
highlighted_rs_ids <- c("rs1548803", "rs3741845", "rs1015442", "rs1015443", 
                        "rs3851584", "rs10772397", "rs1450839", "rs10845279", 
                        "rs10845280", "rs10845281", "rs12226919", "rs12226920", 
                        "rs11054142", "rs11054143", "rs7301234")

highlight_labels <- function(labels) {
  ifelse(labels %in% highlighted_rs_ids, paste0("*", labels, "*"), labels)
}

png("ld_heatmap_chr12.png", width = 25, height = 25, units = "in", res = 450)

col_palette <- colorRampPalette(c("#F0F8FF","#003399"))(200)

n_labels <- nrow(ld_matrix)
gap <- max(1, floor(n_labels / 30))

corrplot(ld_matrix, method = "color", type = "lower", 
         tl.col = ifelse(rownames(ld_matrix) %in% highlighted_rs_ids, "red", "black"), 
         tl.cex = 1.8,  # Adjust label size
         mar = c(0,0,0,0),
         tl.srt = 45,
         diag = FALSE,
         col = col_palette,
         cl.cex = 2.5, 
         addgrid.col = "gray",
         na.label.col = "white",  # Set missing (NA) cells to gray
         is.corr = FALSE,  
         zlim = c(-0.2, 1),  
         cl.ratio = 0.1,
         cl.align = "r",
         gap = gap)

dev.off()

```

Chromosome 7

```{r}
ld_oral <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/independent_markers/r_squared_no_filter.ld", header = T) %>% 
  mutate(R2 = round(R2, 1)) %>% filter(CHR_A == 7)

snp_euro <- read_excel("snp_euro_oral.xlsx", sheet="snp_euro_oral") %>%   
  mutate(label = paste(rs_id, gene, sep = ","))

ld_oral <- ld_oral %>%
  left_join(snp_euro %>% filter(chromosome == 7) %>% select(snp, rs_id), by = c("SNP_A" = "snp")) %>%
  rename(rs_id_A = rs_id) %>%
  left_join(snp_euro %>% filter(chromosome == 7) %>% select(snp, rs_id), by = c("SNP_B" = "snp")) %>%
  rename(rs_id_B = rs_id)

# Create a vector of unique SNP labels
snp_labels <- (snp_euro %>% filter(chromosome == 7))$rs_id

# Create a matrix of R2 values, setting missing pairs as NA
n_snps <- length(snp_labels)
ld_matrix <- matrix(NA, nrow = n_snps, ncol = n_snps)  # Initialize with NA for missing data
rownames(ld_matrix) <- colnames(ld_matrix) <- snp_labels

for (i in 1:nrow(ld_oral)) {
  idx_a <- which(snp_labels == ld_oral$rs_id_A[i])
  idx_b <- which(snp_labels == ld_oral$rs_id_B[i])
  ld_matrix[idx_a, idx_b] <- ld_matrix[idx_b, idx_a] <- ld_oral$R2[i]
}



highlight_labels <- function(labels) {
  ifelse(labels %in% highlighted_rs_ids, paste0("*", labels, "*"), labels)
}

png("ld_heatmap_chr7.png", width = 25, height = 25, units = "in", res = 450)

col_palette <- colorRampPalette(c("#F0F8FF","#003399"))(200)

n_labels <- nrow(ld_matrix)
gap <- max(1, floor(n_labels / 30))

corrplot(ld_matrix, method = "color", type = "lower", 
         tl.col = ifelse(rownames(ld_matrix) %in% highlighted_rs_ids, "red", "black"), 
         tl.cex = 1.8,  # Adjust label size
         mar = c(0,0,0,0),
         tl.srt = 45,
         diag = FALSE,
         col = col_palette,
         cl.cex = 2.5, 
         addgrid.col = "gray",
         na.label.col = "white",  # Set missing (NA) cells to gray
         is.corr = FALSE,  
         zlim = c(-0.2, 1),  
         cl.ratio = 0.1,
         cl.align = "r",
         gap = gap)

dev.off()

```

Chromosome 5

```{r}
ld_oral <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/independent_markers/r_squared_no_filter.ld", header = T) %>% 
  mutate(R2 = round(R2, 1)) %>% filter(CHR_A == 5)

snp_euro <- read_excel("snp_euro_oral.xlsx", sheet="snp_euro_oral") %>%   
  mutate(label = paste(rs_id, gene, sep = ","))

ld_oral <- ld_oral %>%
  left_join(snp_euro %>% filter(chromosome == 5) %>% select(snp, rs_id), by = c("SNP_A" = "snp")) %>%
  rename(rs_id_A = rs_id) %>%
  left_join(snp_euro %>% filter(chromosome == 5) %>% select(snp, rs_id), by = c("SNP_B" = "snp")) %>%
  rename(rs_id_B = rs_id)
snp_labels <- (snp_euro %>% filter(chromosome == 5))$rs_id

# Create a matrix of R2 values, setting missing pairs as NA
n_snps <- length(snp_labels)
ld_matrix <- matrix(NA, nrow = n_snps, ncol = n_snps)  # Initialize with NA for missing data
rownames(ld_matrix) <- colnames(ld_matrix) <- snp_labels

for (i in 1:nrow(ld_oral)) {
  idx_a <- which(snp_labels == ld_oral$rs_id_A[i])
  idx_b <- which(snp_labels == ld_oral$rs_id_B[i])
  ld_matrix[idx_a, idx_b] <- ld_matrix[idx_b, idx_a] <- ld_oral$R2[i]
}


png("ld_heatmap_chr5.png", width = 25, height = 25, units = "in", res = 450)

col_palette <- colorRampPalette(c("#F0F8FF","#003399"))(200)
corrplot(ld_matrix, method = "color", type = "lower", 
         tl.col = ifelse(rownames(ld_matrix) %in% highlighted_rs_ids, "red", "black"),
         tl.cex = 4,  # Adjust label size
         mar = c(0,0,0,0),
         tl.srt = 45,
         diag = FALSE,
         col = col_palette,
         cl.cex = 3.5, 
        na.label.col = "white",  # Set missing (NA) cells to gray
         addgrid.col = "gray")

dev.off()

```

## Classify based on type

```{r}
oral_classify <- snp_euro %>% group_by(gene) %>% tabyl(gene)
snp_euro <- snp_euro %>% mutate(adjusted_p_value = as.numeric(adjusted_p_value))

oral_classify <- left_join(oral_classify, snp_euro %>% filter(adjusted_p_value <0.05) %>% group_by(gene) %>% tabyl(gene) %>% select(-percent) %>% rename(n_different=n), by="gene") %>% 
  mutate(relative_different=n_different/n)
write.csv(oral_classify, "gene_type_classification.csv", col.names = T)
```

Read UKB files

```{r}
ukb_ra <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/UKB/RA", header = TRUE)
ukb_osteo <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/UKB/osteoarthritis", header = TRUE)
ukb_variants <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/UKB/variants.tsv", header = TRUE)

#ra
rsid_list <- c("rs1548803", "rs3741845", "rs1015442", "rs1015443", 
               "rs3851584", "rs10772397", "rs1450839", "rs10845279", 
               "rs10845280", "rs10845281", "rs12226919", "rs12226920", 
               "rs11054142", "rs11054143", "rs7301234")

ukb_ra_rsid <- ukb_ra  %>% 
  left_join(ukb_variants %>% select(variant, rsid, consequence), by = "variant") %>%
  filter(rsid %in% rsid_list)

#osteo
ukb_osteo_rsid <- ukb_osteo  %>% 
  left_join(ukb_variants %>% select(variant, rsid, consequence), by = "variant") %>%
  filter(rsid %in% rsid_list)

write.csv(ukb_ra_rsid, "ukb_ra.csv", col.names = T)
```

```         
```
