---
title: "OralHealth"
format: html
editor: visual
---

## Data preparation

```{r}
snp_euro <- final_clsa_european %>% filter(nchar(major_allele) == 1 & nchar(minor_allele) ==1)
chr <- snp_euro$chromosome
bp <- snp_euro$position
tas2r <- snp_euro$tas2r

write.table(data.frame(chr, bp,bp,tas2r), "snp_euro.txt", quote = F, row.names = F, col.names = F)
```

tas2r snp file for plink (n=124)

```{r}

new_data <- data.frame(snp_euro$chromosome, snp_euro$position, snp_euro$position, snp_euro$tas2r)

write.table(new_data, "snp_euro_all.txt", quote = F, row.names = F, col.names = F)
```

Load baseline data

```{r}
bl_com <- read.csv("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_CoPv7.csv") 
bl_trm <- read.csv("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_Trav4.csv") 
bl_com_dict <- read_excel("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_CoPv7-dictionary.xlsx")
bl_trm_dict <- read_excel("/home/mojan/Mojan_data/CLSA COVID-Baseline/BL/Baseline_Trav4-dictionary.xlsx")
```

Extract oral health variables

```{r}
oral_variables <- bl_com_dict %>% filter(str_detect(label,"Oral health")|str_detect(label, "teeth")|str_detect(label,"eating in")|str_detect(label, "tooth"))%>% select(-table)

#include sex, age, gwas_id, and oral health variables 
oral_health <- bl_com %>%rename(gwas_id= ADM_GWAS3_COM, sex = SEX_ASK_COM, age =AGE_NMBR_COM) %>% select(gwas_id, sex, age, any_of(oral_variables$name))
#write.csv(oral_health, "oral_health.csv")
#write.csv(oral_variables  %>% clean_names(), "oral_variables.csv")
```

## Create phenotype file for plink

```{r}
#Manually changed the names to lables
oral_health <- read.csv("oral_health.csv") %>% clean_names()
#european_id <- european_id %>%   rename(FID=fid, IID=iid) 


#add case-defined variables
oral_health <- oral_health %>%  mutate(periodontal_diseases=case_when(bleeding_gums_around_natural_teeth==1|loose_natural_tooth==1 ~ 1, TRUE ~0)) %>% 
  rename(FID=fid, IID=iid) 
european_id <- european_id %>%  rename(FID=fid, IID=iid) 


#include european ancestry
oral_health_european <- left_join(european_id, oral_health, by=c("FID", "IID")) %>% select(-x, -sex, -age) 



#modify scale, affected=2, unaffected=1, except for has_one_or_more_of_original_tooth
oral_health_binary_euro <- oral_health_european %>%  select(-general_oral_health, -has_one_or_more_of_original_teeth, -uncomfortable_eating_in_last_12_months, -avoided_eating_in_last_12_months, -frequency_brushing_teeth_and_or_dentures, -frequency_brushing_teeth_and_or_dentures_unit_of_measure, -ohp_other_specify_01, -ohp_other_specify_02, -ohp_other_specify_03, -ohp_other_specify_04, -ohp_other_specify_05) %>% mutate_all(.funs = list(~ifelse(. == 1, 2, ifelse(. == 0, 1, .)))) 

#add dental problems case definition
oral_health_binary_euro <- oral_health_binary_euro %>%  mutate(dental_problems=case_when(decayed_natural_tooth==2|toothache==2|broken_natural_tooth==2 ~ 2, decayed_natural_tooth==1|toothache==1|broken_natural_tooth==1 ~ 1, TRUE ~ NA)) %>% #temporomandibular pain
  mutate(temporomandibular_pain=case_when(sore_jaw_muscles==2| painful_jaw_joints==2 ~ 2, sore_jaw_muscles==1|painful_jaw_joints==1 ~ 1, TRUE ~ NA))

#oral_health_status and avoided/uncomfrotable eating binary case definition
oral_health_binary_euro <- left_join(oral_health_binary_euro, 
                                     oral_health_european %>% select(FID, general_oral_health, has_one_or_more_of_original_teeth, avoided_eating_in_last_12_months, uncomfortable_eating_in_last_12_months) %>% mutate(poor_general_oral_health=case_when(general_oral_health==1|general_oral_health==2|general_oral_health==3 ~ 1, TRUE ~ 2)), by="FID") %>% rename(edentulism="has_one_or_more_of_original_teeth") %>% select(-wears_dentures) %>% 
  mutate(avoided_eating=case_when(avoided_eating_in_last_12_months==1| avoided_eating_in_last_12_months==2 ~ 2, TRUE ~1)) %>% 
  mutate(uncomfortable_eating=case_when(uncomfortable_eating_in_last_12_months==1| uncomfortable_eating_in_last_12_months==2 ~ 2, TRUE ~1)) 

oral_health_binary_euro$edentulism[oral_health_binary_euro$edentulism == 8] <- NA

#avoided_eating_binary case definition

 
#Remove individuals with edentulism + reported at least one problem related to tooth
oral_health_binary_euro <- oral_health_binary_euro %>%
  filter(!(edentulism == 2 & (
    toothache == 2 |
    decayed_natural_tooth == 2 |
    loose_natural_tooth == 2 |
    broken_natural_tooth == 2 |
    sore_gums_around_natural_teeth == 2 |
    bleeding_gums_around_natural_teeth == 2
  )))%>% select(-avoided_eating_in_last_12_months, -general_oral_health, -uncomfortable_eating_in_last_12_months)

oral_health_european <- oral_health_european %>%
  filter(!(has_one_or_more_of_original_teeth == 2 & (
   toothache == 1 |
    decayed_natural_tooth == 1 |
    loose_natural_tooth == 1 |
    broken_natural_tooth == 1 |
    sore_gums_around_natural_teeth == 1 |
    bleeding_gums_around_natural_teeth == 1
  ))) %>% select(-avoided_eating_in_last_12_months, -general_oral_health, -uncomfortable_eating_in_last_12_months)
write.table(oral_health_binary_euro,"oral_health_binary_euro.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")

```

Create phenotype for for noneuropean for plink not changed for the variable names

```{r}

#include noneuropean ancestry
#noneuropean_id <- noneuropean_id %>% rename(FID=fid, IID=iid)
oral_health_noneuropean <- left_join(noneuropean_id, oral_health, by=c("FID", "IID")) %>% select(-x, -sex, -age)

#modify scale, affected=2, unaffected=1, except for has_one_or_more_of_original_tooth
oral_health_binary_noneuro <- oral_health_noneuropean %>%  select(-general_oral_health, -has_one_or_more_of_original_teeth, -uncomfortable_eating_in_last_12_months, -avoided_eating_in_last_12_months, -frequency_brushing_teeth_and_or_dentures, -frequency_brushing_teeth_and_or_dentures_unit_of_measure, -ohp_other_specify_01, -ohp_other_specify_02, -ohp_other_specify_03, -ohp_other_specify_04, -ohp_other_specify_05) %>% mutate_all(.funs = list(~ifelse(. == 1, 2, ifelse(. == 0, 1, .)))) 


#oral_health_status and avoided/uncomfrotable eating binary case definition
oral_health_binary_noneuro <- left_join(oral_health_binary_noneuro, 
                                     oral_health_noneuropean %>% select(FID, general_oral_health, has_one_or_more_of_original_teeth, avoided_eating_in_last_12_months, uncomfortable_eating_in_last_12_months) %>% mutate(poor_general_oral_health=case_when(general_oral_health==1|general_oral_health==2|general_oral_health==3 ~ 1, TRUE ~ 2)), by="FID") %>% rename(edentulism="has_one_or_more_of_original_teeth") %>% select(-wears_dentures) %>% 
  mutate(avoided_eating=case_when(avoided_eating_in_last_12_months==1| avoided_eating_in_last_12_months==2 ~ 2, TRUE ~1)) %>% 
  mutate(uncomfortable_eating=case_when(uncomfortable_eating_in_last_12_months==1| uncomfortable_eating_in_last_12_months==2 ~ 2, TRUE ~1)) 

oral_health_binary_noneuro$edentulism[oral_health_binary_noneuro$edentulism == 8] <- NA


#Remove individuals with edentulism + reported at least one problem related to tooth
oral_health_binary_noneuro <- oral_health_binary_noneuro %>%
  filter(!(edentulism == 2 & (
    toothache == 2 |
   decayed_natural_tooth == 2 |
    loose_natural_tooth == 2 |
    broken_natural_tooth == 2 |
    sore_gums_around_natural_teeth == 2 |
    bleeding_gums_around_natural_teeth == 2
  )))

oral_health_binary_euro <- inner_join(oral_health_binary_euro, oral_prevalence %>% select(IID, Sex, Age, Smoking_status, Brushing_frequency), by="IID")

write.table(oral_health_binary_noneuro,"oral_health_binary_noneuro.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")

```

## Chi-squared test with plink

```{bash}

OUT=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output

#extract list of unique tas2r snps in european population with available 1000 Genomes european allele frequency, produced by R
plink --bfile $OUT/clsa_imp_v3_european_sm_qc_maf_0.01 --extract $OUT/independent_markers/list_eur_snps_to_extract_independent_snps.txt range --make-bed --out $OUT/independent_markers/snps_with_1kgp_af_european_sm_qc_maf0.01


#independent markers

plink --bfile $OUT/independent_markers/snps_with_1kgp_af_european_sm_qc_maf0.01 --indep-pairwise 1000 kb 5 0.1 --out $OUT/independent_markers/independent_snps_with_1kgp_af_european_sm_qc_maf0.01

#independent markers in n=124 snps
plink --bfile $OUT/snp_t2r_european_sm_qc_maf0.01 --indep-pairwise 1000 kb 5 0.1 --out $OUT/independent_markers/independent_snps_t2r_european_sm_qc_maf0.01


#calculate r-squared for linkage disequilibrium between the SNPs n=124

plink --bfile $OUT/snp_t2r_european_sm_qc_maf0.01 --r2 --ld-window 99999 --ld-window-kb 1000 --ld-window-r2 0.2 --out $OUT/independent_markers/r_squared
```

## Summarize plink chi-squared outcomes

```{r}
final_clsa_european <- left_join(clsa_european, consequence_filtered_european %>% select(snp, rs_id, symbol, consequence, eur_af, p_value, protein_position, amino_acids, sift, poly_phen, strand, type, id), by="snp")


phenotypes <- c(
  "toothache",
  "cannot_chew_adequately",
  "swelling_in_mouth",
  "dry_mouth",
  "burning_mouth",
  "sore_jaw_muscles",
  "painful_jaw_joints",
  "decayed_natural_tooth",
  "loose_natural_tooth",
  "broken_natural_tooth",
  "sore_gums_around_natural_teeth",
  "bleeding_gums_around_natural_teeth",
  "denture_related_sores",
  "dirty_teeth_or_dentures",
  "bad_breath",
  "periodontal_diseases",
  "edentulism",
  "poor_general_oral_health",
  "avoided_eating",
  "uncomfortable_eating",
  "no_oral_health_problems",
  "dental_problems",
  "temporomandibular_pain"
)
```

```{r}
gwas_outcomes_oral <- list()
gwas_combined <- read.table("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/univariable_results/gwas.toothache.assoc",header = TRUE) %>% 
  clean_names %>% select(snp, p) %>% mutate_all(tolower)%>% left_join(., snp_euro %>% select(snp, gene, rsid, consequence), by="snp")

gwas_combined_significant <- read.table("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/univariable_results/gwas.toothache.assoc",header = TRUE) %>% 
  clean_names %>% select(snp, p) %>% mutate_all(tolower) %>% left_join(., snp_euro, by="snp")

# L0.002088oop through each phenotype
for (phenotype in phenotypes) {
  # Read the corresponding file
  file_path <- paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/univariable_results/gwas.", phenotype, ".assoc")
  data <- read.table(file_path, header = TRUE)%>% clean_names() %>%
    mutate_all(tolower) %>% mutate(adjusted_p=as.numeric(p)*35)
  
  # Assign the dataframe to the list with the same name as phenotype
  gwas_outcomes_oral[[phenotype]] <- left_join(data, snp_euro %>% select(snp, gene, rsid, consequence), by="snp")
  
    gwas_combined_significant <- left_join(gwas_combined_significant, data %>% filter(adjusted_p<0.05)%>% select(snp, adjusted_p, or) %>% 
                               rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, !!paste0("or_", phenotype) := or), by="snp")

  
  gwas_combined <- left_join(gwas_combined, data %>% 
                               select(snp, adjusted_p, or) %>% 
                               rename(!!paste0("adjusted_p_", phenotype) := adjusted_p, !!paste0("or_", phenotype) := or), 
                             by = "snp")
  
}
gwas_combined_significant <- gwas_combined_significant[rowSums(!is.na(gwas_combined_significant %>% select(-snp, -p, -gene, -rsid, -consequence))) > 0, ]



write.csv(gwas_combined, "gwas_combined.csv", col.names = TRUE)
write.csv(gwas_combined_significant, "gwas_combined_significant.csv", col.names = TRUE)
```

## Prevalence of outcomes

```{r}

#used oral_health_european because also includes categorical variables of multiple categories
oral_health_european <- left_join(oral_health_european, bl_com %>% rename(IID= ADM_GWAS3_COM, sex = SEX_ASK_COM, age =AGE_NMBR_COM) %>%  select(IID, sex, age), by= "IID") %>% mutate(sex=case_when(sex=="M"~1, sex=="F"~2))


oral_health_european %>%  tbl_summary(include = everything(), by=sex, label = list(oral_health_status ~ "General Oral Health"),
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary.docx") 

#gt summary for the total european(not by sex)
oral_health_binary_euro %>%  tbl_summary(include = everything(), 
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary_european.docx") 

oral_health_binary_euro %>%  tbl_summary(include = everything(), by=Sex,
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary_european_by_sex.docx") 

#gt summary for the total noneuropean
oral_health_binary_noneuro %>%  tbl_summary(include = everything(),
    missing_text = "(Missing)") %>%
 as_gt() %>%
  gt::gtsave(filename = "oral_health_summary_non_european.docx") 

```

Plot prevalences

```{r}

oral_health_sum <- read.csv("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/oral_health_summary.csv")
oral_health_sum$Characteristic <- factor(oral_health_sum$Characteristic, levels = unique(oral_health_sum$Characteristic))
oral_health_long <- gather(oral_health_sum, key = "Gender", value = "Percentage", Males, Females)

plot <- ggplot(oral_health_long, aes(x = Characteristic, y = Percentage, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Prevalence of Oral Health Symptoms", x = "Characteristic", y = "Percentage (%)") +
  scale_fill_manual(values = c("Males" = "cyan", "Females" = "magenta"))+
  theme(text = element_text(size = 25))
ggsave("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/oral_health_sum.tiff", 
       plot, height = 12*0.75, width = 15, units = 'in', dpi = 120)


```

## Manhattan plot for chi-squared test results

```{r}
# Set the size of the graphic device
dev.new(width=11, height=4.5)

for (phenotype in phenotypes) {

# Prepare the data
gwas_results <- read.table(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/univariable_results/gwas.", phenotype, ".assoc"), header = TRUE) %>% clean_names() %>% rename(chromosome=chr) %>%
  mutate_all(tolower) %>% mutate(p=as.numeric(p)) %>% mutate(chromosome=as.numeric(chromosome)) 

gwas_results <- inner_join(gwas_results, snp_euro, by=c("chromosome", "snp")) 

gwas_results <- gwas_results %>%
  mutate(log_p = -log10(p)) %>% 
  arrange(as.numeric(chromosome), bp) %>% 
  mutate(label = paste(snp, tas2r, consequence, sep = "_"))  # create new LABEL column

# Create an ordered factor for GENE based on the order in the arranged dataframe
gwas_results$tas2r <- factor(gwas_results$tas2r, levels = unique(gwas_results$tas2r)) 

# Set a color palette with a large number of colors
num_colors <- length(unique(gwas_results$tas2r))
my_palette <- rainbow(num_colors)

# Shuffle the palette
set.seed(123)  # for reproducible results
my_palette <- sample(my_palette)

# Create a scatter plot
manhattan_plot = ggplot(gwas_results, aes(x = tas2r, y = log_p, color = tas2r)) +
  geom_point(alpha = 0.5, size=3.5) +
  scale_color_manual(values = my_palette) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 2.8, linetype = "dashed", color = "red") +
  geom_text(data = subset(gwas_results, log_p > 3.5), aes(label = label), size = 5, vjust = 0, hjust = 0,check_overlap = F) +
  # ggrepel::geom_text_repel(data = subset(gwas_results, log_p_bonferroni > 4), aes(label = LABEL), 
  #                        nudge_x = 0., nudge_y = 0.0, size = 3, check_overlap = F) +
  labs(title = gsub("_|ohp", " ", phenotype), x = NULL, y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 30), plot.title = element_text(size = 30, face = "bold")) +
  guides(color = FALSE) +
  scale_y_continuous(limits = c(0, 3.2))
# Save the plot
ggsave(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/chi_squared_manhattan_plots/", phenotype, ".png"), plot = manhattan_plot, width = 11, height = 4.5, dpi = 450)
}
```

compare males with females

```{r}
df <- data.frame(
  Characteristic = c(
    "Smoking_status: 1", "Smoking_status: 2", "Smoking_status: 3",
    "General oral health: Excellent", "General oral health: Very good", "General oral health: Good", 
    "General oral health: Fair", "General oral health: Poor",
    "Uncomfortable eating in last 12 months: Often", "Uncomfortable eating in last 12 months: Sometimes", 
    "Uncomfortable eating in last 12 months: Rarely", "Uncomfortable eating in last 12 months: Never",
    "Avoided eating in last 12 months: Often", "Avoided eating in last 12 months: Sometimes", 
    "Avoided eating in last 12 months: Rarely", "Avoided eating in last 12 months: Never",
    "Edentulism", "Toothache", "Cannot chew adequately", "Swelling in mouth", 
    "Dry mouth", "Burning mouth", "Jaw muscles sore", "Jaw joints painful", 
    "Natural tooth decayed", "Natural tooth loose", "Natural tooth broken", 
    "Gums around natural teeth are sore", "Gums around natural teeth bleed", 
    "Denture related sores", "Teeth or dentures dirty", "Bad breath", 
    "Periodontal disease", "None", "Other", "No answer"
  ),
  Males = c(
    859, 5553, 4659, 
    3414, 4407, 2399, 545, 141,
    193, 717, 2135, 7857,
    143, 399, 1031, 9330,
    629, 1410, 799, 465,
    1689, 102, 396, 404,
    1798, 510, 1278, 754, 1094,
    203, 386, 823, 1508,
    4754, 365, 4
  ),
  Females = c(
    1013, 4816, 5089, 
    3958, 4318, 2210, 447, 120,
    296, 931, 2197, 7631,
    213, 634, 1313, 8893,
    686, 1306, 933, 535,
    2196, 207, 788, 868,
    1352, 522, 1124, 894, 1315,
    262, 239, 735, 1700,
    4618, 444, 16
  )
)

# Function to compute p-value
compute_p_value <- function(males, females) {
  test <- chisq.test(matrix(c(males, females), ncol = 2))
  return(test$p.value)
}

# Apply the function to each row and add a new column with p-values
df <- df %>%
  rowwise() %>%
  mutate(p_value = compute_p_value(Males, Females))
write.csv(df, "~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/summary_compare_sex.csv")

```

compare specific conditions

```{r}
# Data for the contingency table
data <- data.frame(
  Condition = c("Periodontal disease", "Dental problems", "Temporomandibular pain"),
  Males = c(1508, 3298 , 610),
  Females = c(1700, 2831 , 1198)
)

# Calculate p-values using chi-squared test
p_values <- apply(data[, 2:3], 1, function(x) chisq.test(x)$p.value)

# Add p-values to the table
data$P_Value <- round(p_values, 6)

# Display the table
print(data)

```

## Create covariate files for plink

```{r}
covariates_oral_list <- read_excel("oral health covariates.xlsx")
#Sex 1=male; 2=female; other=unknown
covariates_oral <- bl_com %>% select(ADM_GWAS3_COM,AGE_NMBR_COM, SEX_ASK_COM, ICQ_SMOKE_COM, ORH_BRUSH_NB_MCQ, ORH_BRUSH_UNIT_MCQ) %>%
  rename(IID=ADM_GWAS3_COM)
covariates_oral <- left_join(european_id, covariates_oral, by="IID") %>% mutate(FID=IID) %>% mutate(SEX_ASK_COM=case_when(SEX_ASK_COM=="M" ~ 1, SEX_ASK_COM=="F" ~ 2, TRUE ~ 0))
#convert brushing value
covariates_oral <- covariates_oral %>% mutate(brushing_frequency=case_when(ORH_BRUSH_UNIT_MCQ==1 ~ ORH_BRUSH_NB_MCQ*365,
                                                                                    ORH_BRUSH_UNIT_MCQ==2 ~ ORH_BRUSH_NB_MCQ*52,
                                                                                    ORH_BRUSH_UNIT_MCQ==3 ~ ORH_BRUSH_NB_MCQ*12,
                                                                                    ORH_BRUSH_UNIT_MCQ==4 ~ ORH_BRUSH_NB_MCQ,
                                                                                      TRUE ~ NA)) %>% 
  rename(Age=AGE_NMBR_COM, Sex=SEX_ASK_COM, Brushing_Frequency = brushing_frequency, Smoking_Status = ICQ_SMOKE_COM) %>% 
  select(-ORH_BRUSH_NB_MCQ, -ORH_BRUSH_UNIT_MCQ)
write.table(covariates_oral, "covariates_oral.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")

write.table(covariates_oral %>% select(FID, IID, Sex), "update_sex.txt", quote = FALSE, row.names = FALSE,
            sep = "\t")
```

Covariates for european

```{r}
#Sex 1=male; 2=female; other=unknown
covariates_oral <- bl_com %>% select(ADM_GWAS3_COM,AGE_NMBR_COM, SEX_ASK_COM, ICQ_SMOKE_COM, ORH_BRUSH_NB_MCQ, ORH_BRUSH_UNIT_MCQ) %>%
  rename(IID=ADM_GWAS3_COM)

covariates_oral <- covariates_oral %>% #rename(IID=ADM_GWAS_COM) 
  mutate(FID=IID) 
#covariates_oral_noneuro <- left_join(noneuropean %>% select(IID, pca.cluster.id), covariates_oral, by="IID")

#%>% mutate(SEX_ASK_COM=case_when(SEX_ASK_COM=="M" ~ 1, SEX_ASK_COM=="F" ~ 2, TRUE ~ 0))
#convert brushing value
covariates_oral <- covariates_oral %>% mutate(brushing_frequency=case_when(ORH_BRUSH_UNIT_MCQ==1 ~ ORH_BRUSH_NB_MCQ*365,
                                                                                    ORH_BRUSH_UNIT_MCQ==2 ~ ORH_BRUSH_NB_MCQ*52,
                                                                                    ORH_BRUSH_UNIT_MCQ==3 ~ ORH_BRUSH_NB_MCQ*12,
                                                                                    ORH_BRUSH_UNIT_MCQ==4 ~ ORH_BRUSH_NB_MCQ,
                                                                                      TRUE ~ NA)) %>% 
  rename(Age=AGE_NMBR_COM, Sex=SEX_ASK_COM, Brushing_frequency = brushing_frequency, Smoking_status = ICQ_SMOKE_COM) %>% 
  select(-ORH_BRUSH_NB_MCQ, -ORH_BRUSH_UNIT_MCQ) %>% select(FID, IID, everything())


#write.table(covariates_oral_noneuro, "covariates_oral_noneuro.txt", quote = FALSE, row.names = FALSE,  sep = "\t")

```

## Assess multicollinearity

```{r}
library(corrplot)
corrplot(cor(covariates_oral %>% select(-IID), use="complete.obs"), method="circle", # Increase space for labels
         type="lower", tl.srt=45, mar.labels = c(2,2))

corrplot(cor(covariates_oral_noneuro %>% select(-IID, -FID), use="complete.obs"), method="circle", # Increase space for labels
         type="lower", tl.srt=45, mar.labels = c(2,2))

```

## Logistic regression with PLINK

```{bash}
DATA=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output

#logistic regression

plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex\
          --logistic --pheno $DATA/oral_health/oral_health_binary_euro.txt --all-pheno\
          --covar $DATA/oral_health/covariates_oral.txt \
          --covar-name Age, Smoking_status, Brushing_frequency, Sex\
          --out $DATA/oral_health/logistic_results/gwas
          
#plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex --recode --out $DATA/oral_health/snp_t2r_european_sm_qc_maf0.01_updated_sex

```

## Summarize logistic regression results

```{r}
logistic_results_oral <- list()
logistic_combined_oral <- read.table("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.toothache.assoc.logistic",header = TRUE) %>% 
  clean_names %>% mutate_all(tolower) %>% filter(test=="add")  %>%  left_join(., snp_euro, by="snp")

logistic_combined_oral_significant <- read.table("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.toothache.assoc.logistic",header = TRUE) %>% clean_names %>% mutate_all(tolower) %>% filter(test=="add")  %>%  left_join(., snp_euro, by="snp") %>% select(-p, -or, -test, -stat)


# Loop through each phenotype
for (phenotype in phenotypes) {
  # Read the corresponding file
  file_path <- paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic")
  data <- read.table(file_path, header = TRUE)%>% clean_names() %>%
    mutate_all(tolower) %>% filter(test=="add")
  
  # Assign the dataframe to the list with the same name as phenotype
  logistic_results_oral[[phenotype]] <- left_join(data, snp_euro %>% select(snp, gene, rsid, consequence), by="snp")
  
   logistic_combined_oral_significant <- left_join(logistic_combined_oral_significant, data %>% filter(p<0.05)%>% select(snp, p, or) %>% 
                               rename(!!paste0("p_", phenotype) := p, !!paste0("or_", phenotype) := or), by="snp")
  
  logistic_combined_oral <- left_join(logistic_combined_oral, data %>% 
                               select(snp, p, or) %>% 
                               rename(!!paste0("p_", phenotype) := p, !!paste0("or_", phenotype) := or), 
                             by = "snp")
  
}

 #logistic_combined_oral_significant <-  logistic_combined_oral_significant[rowSums(!is.na(
   #logistic_combined_oral_significant%>% select(-snp, -p, -tas2r, -rs_id, -symbol, -consequence))) > 0, ]

write.csv(logistic_combined_oral, "logistic_combined_oral.csv", col.names = TRUE)
write.csv(logistic_combined_oral_significant, "logistic_combined_oral_significant.csv", col.names = TRUE)

#write.csv(inner_join(read_excel("logistic_combined_oral_significant.xlsx", sheet="t2r4_14_20"), snp_euro, by="snp"), "logistic_combined_oral_significant.csv", col.names = TRUE)
```

Manhattan plots

```{r}
# Set the size of the graphic device
dev.new(width=11, height=4.5)

for (phenotype in phenotypes) {

# Prepare the data
gwas_results <- read.table(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic"), header = TRUE) %>% clean_names() %>% rename(chromosome=chr) %>%
  mutate_all(tolower) %>% mutate(p=as.numeric(p)) %>% mutate(chromosome=as.numeric(chromosome)) %>% filter(test=="add")

gwas_results <- inner_join(gwas_results, snp_euro, by=c("snp"))

gwas_results <- gwas_results %>%
  mutate(log_p = -log10(p)) %>% 
  arrange(as.numeric(chromosome), bp) %>% 
  mutate(label = paste(snp, gene, consequence, sep = "_"))  # create new LABEL column

# Create an ordered factor for GENE based on the order in the arranged dataframe
gwas_results$tas2r <- factor(gwas_results$gene, levels = unique(gwas_results$tas2r))

# Set a color palette with a large number of colors
num_colors <- length(unique(gwas_results$gene))
my_palette <- rainbow(num_colors)

# Shuffle the palette
set.seed(123)  # for reproducible results
my_palette <- sample(my_palette)

# Create a scatter plot
manhattan_plot = ggplot(gwas_results, aes(x = tas2r, y = log_p, color = tas2r)) +
  geom_point(alpha = 0.5, size = 3.5) +
  scale_color_manual(values = my_palette) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  geom_text(data = subset(gwas_results, log_p > 3.5), aes(label = label), size = 6, vjust = 0, hjust = 0,check_overlap = F) +
  # ggrepel::geom_text_repel(data = subset(gwas_results, log_p_bonferroni > 4), aes(label = LABEL), 
  #                        nudge_x = 0., nudge_y = 0.0, size = 3, check_overlap = F) +
  labs(title = gsub("_|ohp", " ", phenotype), x = NULL, y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 30),  plot.title = element_text(size = 30, face = "bold")) +
  guides(color = FALSE) +
  scale_y_continuous(limits = c(0, 3))

# Save the plot
ggsave(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logitmanhattan_plots/", phenotype, ".png"), plot = manhattan_plot, width = 11, height = 4.5, dpi = 450)
}
```

TAS2R20

```{r}
all_gwas_tas2r20 <- data.frame()

# Loop through phenotypes to gather the data
for (phenotype in phenotypes) {
  
  gwas_results <- read.table(paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/logistic_results/gwas.", phenotype, ".assoc.logistic"), header = TRUE) %>% 
    clean_names() %>% 
    rename(chromosome = chr) %>%
    mutate_all(tolower) %>% 
    mutate(p = as.numeric(p)) %>% 
    mutate(chromosome = as.numeric(chromosome)) %>% filter(test=="add")
  
  gwas_results <- inner_join(gwas_results, snp_euro, by = c("chromosome", "snp")) %>%
    mutate(log_p = -log10(p)) %>% 
    arrange(as.numeric(chromosome), bp) %>% 
    mutate(label = paste(snp, tas2r, consequence, sep = "_"))

  gwas_tas2r20 <- gwas_results %>% 
    filter(tas2r == "tas2r20") %>%
    mutate(phenotype = phenotype)
  
  all_gwas_tas2r20 <- bind_rows(all_gwas_tas2r20, gwas_tas2r20)
}

# Create a scatter plot
manhattan_plot <- ggplot(all_gwas_tas2r20, aes(x = phenotype, y = log_p, color = rs_id)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "blue") +
  labs(y = "-log10(p-value)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), text = element_text(size = 27),  plot.title = element_text(size = 22, face = "bold")) +
  scale_y_continuous(limits = c(0, 2.5))

# Save the plot
ggsave("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/tas2r20/combined_plot.png", plot = manhattan_plot, width = 10, height = 6, dpi = 450)

```
