---
title: "PRS"
format: html
editor: visual 
---

## First approach: Base=GWAS results from UKB, Target=CLSA European population

Create base data for the target phenotypes from ukb GWAS data

Extract tas2r SNPs

```{r}
ukb_sore_gums <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/prs/uk_biobank/6149_2.gwas.imputed_v3.both_sexes.tsv", header = T)
ukb_bleeding_gums <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/prs/uk_biobank/6149_3.gwas.imputed_v3.both_sexes.tsv", header = T)
ukb_loose_teeth <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/prs/uk_biobank/6149_4.gwas.imputed_v3.both_sexes.tsv", header = T)
ukb_toothache <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/prs/uk_biobank/6149_5.gwas.imputed_v3.both_sexes.tsv", header = T)
```

Merge clsa snp euro with ukb snp variation file (ukb based on GRCH37, so cannot merge with the other files with position information, only with the rsid)

```{r}
ukb_variants <- read.table("/home/mojan/Mojan_data/Genomics3_clsa/output/oral_health/prs/uk_biobank/variants.tsv", header = T)

snp_euro <- read_excel("snp_euro_oral.xlsx", sheet="snp_euro_oral") %>% clean_names() %>% 
  rename(rsid=rs_id)
t2r_ukb <- inner_join(ukb_variants, snp_euro, by="rsid")
t2r_ukb <- t2r_ukb %>% mutate(minor_allele.y=toupper(minor_allele.y)) 
t2r_clsa_ukb <- t2r_ukb %>% filter(minor_allele.x==minor_allele.y) %>% select(-minor_allele.x) %>% rename(minor_allele=minor_allele.y, BP=pos, CHR=chr) %>% select(rsid, gene, snp) %>% rename(rs_id=rsid, SNP=snp) 

t2r_clsa_ukb <- t2r_clsa_ukb %>%
  # Split the SNP column into four new columns
  separate(SNP, into = c("chr", "pos", "ref", "alt"), sep = ":") %>%
  # Convert the ref and alt columns to uppercase
  mutate(ref = toupper(ref), alt = toupper(alt)) %>%
  # Combine the columns back into the SNP column
  unite("SNP", chr, pos, ref, alt, sep = ":") 

t2r_clsa_ukb <- t2r_clsa_ukb %>% rename(rsid=rs_id)

sore_gums <- inner_join(ukb_sore_gums, ukb_variants %>% select(variant, rsid), by="variant")
bleeding_gums <- inner_join(ukb_bleeding_gums, ukb_variants %>% select(variant, rsid), by="variant")
loose_teeth <- inner_join(ukb_loose_teeth, ukb_variants %>% select(variant, rsid), by="variant")
toothache <- inner_join(ukb_toothache, ukb_variants %>% select(variant, rsid), by="variant")


sore_gums <- inner_join(sore_gums, t2r_clsa_ukb, by="rsid")
bleeding_gums <- inner_join(bleeding_gums, t2r_clsa_ukb, by="rsid") 
loose_teeth <- inner_join(loose_teeth, t2r_clsa_ukb, by="rsid")
toothache <- inner_join(toothache, t2r_clsa_ukb, by="rsid") 

```

Merge base information with rsid from the variation file

```{r}
sore_gums <- sore_gums %>% 
  mutate(OR = exp(beta)) %>% 
  rename(N=n_complete_samples, SE=se, MAF=minor_AF, P=pval, A1=minor_allele, rs_id=rsid) %>% 
  select(SNP, A1, N, SE, P, OR, MAF) %>% separate(SNP, into = c("chr", "pos", "ref", "alt"), sep = ":") %>%
  # Convert the ref and alt columns to uppercase
  mutate(ref = toupper(ref), alt = toupper(alt)) %>%
  # Combine the columns back into the SNP column
  unite("SNP", chr, pos, ref, alt, sep = ":") 


bleeding_gums <- bleeding_gums %>% 
  mutate(OR = exp(beta)) %>% 
  rename(N=n_complete_samples, SE=se, MAF=minor_AF, P=pval, A1=minor_allele, rs_id=rsid) %>% 
  select(SNP, A1, N, SE, P, OR, MAF) %>% 
  separate(SNP, into = c("chr", "pos", "ref", "alt"), sep = ":") %>%
  # Convert the ref and alt columns to uppercase
  mutate(ref = toupper(ref), alt = toupper(alt)) %>%
  # Combine the columns back into the SNP column
  unite("SNP", chr, pos, ref, alt, sep = ":") 

loose_teeth <- loose_teeth %>% 
  mutate(OR = exp(beta)) %>% 
  rename(N=n_complete_samples, SE=se, MAF=minor_AF, P=pval, A1=minor_allele, rs_id=rsid) %>% 
  select(SNP, A1, N, SE, P, OR, MAF) %>% 
  separate(SNP, into = c("chr", "pos", "ref", "alt"), sep = ":") %>%
  # Convert the ref and alt columns to uppercase
  mutate(ref = toupper(ref), alt = toupper(alt)) %>%
  # Combine the columns back into the SNP column
  unite("SNP", chr, pos, ref, alt, sep = ":") 


toothache <- toothache %>% 
  mutate(OR = exp(beta)) %>% 
  rename(N=n_complete_samples, SE=se, MAF=minor_AF, P=pval, A1=minor_allele, rs_id=rsid) %>% 
  select(SNP, A1, N, SE, P, OR, MAF) %>% 
  separate(SNP, into = c("chr", "pos", "ref", "alt"), sep = ":") %>%
  # Convert the ref and alt columns to uppercase
  mutate(ref = toupper(ref), alt = toupper(alt)) %>%
  # Combine the columns back into the SNP column
  unite("SNP", chr, pos, ref, alt, sep = ":") 
```

Create base data file for the four phenotypes

```{r}
write.table(bleeding_gums, "~/Mojan_data/Genomics3_clsa/output/oral_health/prs/base_ukb/ukb_bleeding_gums.txt", row.names = FALSE, quote = FALSE, sep = "\t")
write.table(sore_gums, "~/Mojan_data/Genomics3_clsa/output/oral_health/prs/base_ukb/ukb_sore_gums.txt", row.names = FALSE, quote = FALSE, sep = "\t")
write.table(loose_teeth, "~/Mojan_data/Genomics3_clsa/output/oral_health/prs/base_ukb/ukb_loose_teeth.txt", row.names = FALSE, quote = FALSE, sep = "\t")
write.table(toothache, "~/Mojan_data/Genomics3_clsa/output/oral_health/prs/base_ukb/ukb_toothache.txt", row.names = FALSE, quote = FALSE, sep = "\t")


```

Create target data and apply PRSice package

```{bash}
DATA=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs
OUT=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output


#1. create target data (binary files of the eruopean population)
plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex --keep $DATA/oral_prs_target_population.txt\
      --maf 0.01 --geno 0.05 --hwe 0.000000000315 --mind 0.05 \
      --make-bed --out $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex_target
      
PHENO=(
   "bleeding_gums" "loose_teeth" "sore_gums" "toothache" 
)

# 2. prs
  for pheno in "${PHENO[@]}"
  do
    Rscript $DATA/PRSice.R \
      --prsice PRSice_linux \
      --base /home/mojan/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/base_ukb/ukb_${pheno}.txt \
      --target $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex_target \
      --binary-target T \
      --pheno $DATA/target_phenotypes_total_euro/${pheno}.txt \
      --cov $DATA/covariates_oral.txt \
      --stat OR \
      --or \
      --out $DATA/outcome_base_ukb/prs_${pheno}
  done
```

combine prs summary files (base=ukb, target=clsa euro)

```{r}
  
  outcome_oral_prs_base_ukb <- data.frame()
phenotypes_ukb <- c("bleeding_gums", "sore_gums", "toothache", "loose_teeth")

for (phenotype in phenotypes_ukb) {
    file_path <- paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/outcome_base_ukb/prs_", phenotype, ".summary")
    data <- read.table(file_path, header = TRUE)  # assuming the file has headers
    data$Phenotype <- phenotype  # adding a Phenotype column
    outcome_oral_prs_base_ukb <- rbind(outcome_oral_prs_base_ukb, data)
}
write.csv(outcome_oral_prs_base_ukb, "~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/outcome_oral_prs_base_ukb.csv")

```

## Second approach: Base=CLSA European population, Target=CLSA nonEuropean population

create target phenotypes file for the whole european cohort

```{r}
phenotypes <- c(
  "toothache",
  "cannot_chew_adequately",
  "swelling_in_mouth",
  "dry_mouth",
  "burning_mouth",
  "sore_jaw_muscles",
  "painful_jaw_joints",
  "decayed_natural_tooth",
  "loose_natural_tooth",
  "broken_natural_tooth",
  "sore_gums_around_natural_teeth",
  "bleeding_gums_around_natural_teeth",
  "denture_related_sores",
  "dirty_teeth_or_dentures",
  "bad_breath",
  "periodontal_diseases",
  "edentulism",
  "poor_general_oral_health",
  "avoided_eating",
  "uncomfortable_eating",
  "no_oral_health_problems",
  "dental_caries",
  "temporomandibular_pain"
)

for (phenotype in phenotypes) {
  write.table(oral_health_binary_euro %>% select(FID, IID, phenotype),
              paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/target_phenotypes_total_euro/", phenotype, ".txt"),
              row.names = FALSE, quote = FALSE)
}

```

Create target phenotype for noneuropean population

```{r}
for (phenotype in phenotypes) {
  write.table(oral_health_binary_noneuro %>% select(FID, IID, phenotype),
              paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/target_noneuropean/", phenotype, ".txt"),
              row.names = FALSE, quote = FALSE)
}
```

Extract TAS2R SNPs Noneuropean population

```{bash}
IMP=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/imp.new
OUT=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output


#apply sample marker qc, total population
plink --bfile $IMP/clsa_imp_merged_v3 --keep $IMP/01_noneuropean_hq_sample.txt\
      --exclude $IMP/02_total_markers_exclude_mcq_rsq.txt\
      --geno 0.05 --hwe 0.000000000315 --mind 0.05 \
      --make-bed --out $OUT/clsa_imp_v3_noneuropean_sm_qc


#Extract SNPs only (indels were removed using R)

plink --bfile $OUT/clsa_imp_v3_noneuropean_sm_qc --extract $OUT/snp_euro.txt range --make-bed --out $OUT/snp_t2r_noneuropean_sm_qc_maf0.01


plink --bfile $OUT/snp_t2r_noneuropean_sm_qc_maf0.01 --update-sex $OUT/oral_health/update_sex_noneuropean.txt --make-bed --out $OUT/snp_t2r_noneuropean_sm_qc_maf0.01_updated_sex
```

## PRSice package

```{bash}
DATA=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs
OUT=/home/mojan/Mojan_data/Genetics/Genomics3_clsa/output



#create base data (gwas results of european population)
plink --bfile $DATA/snp_t2r_european_sm_qc_maf0.01_updated_sex \
          --pheno $DATA/oral_health_binary_euro.txt \
          --maf 0.01 --geno 0.05 --hwe 0.000000000315 --mind 0.05 \
          --assoc --all-pheno\
          --out $DATA/base_gwas/gwas
          
          
          
#create target data (binary files of noneuropean population)
plink --bfile $OUT/snp_t2r_noneuropean_sm_qc_maf0.01_updated_sex \
      --maf 0.01 --geno 0.05 --hwe 0.000000000315 --mind 0.05 \
      --make-bed --out $DATA/snp_t2r_noneuropean_sm_qc_maf0.01_updated_sex_target
      
      #calculate PRS using PRSice package 

PHENO=(
  "toothache" "cannot_chew_adequately" "swelling_in_mouth" "dry_mouth"
  "burning_mouth"
  "sore_jaw_muscles"
  "painful_jaw_joints"
  "decayed_natural_tooth"
  "loose_natural_tooth"
  "broken_natural_tooth"
  "sore_gums_around_natural_teeth"
  "bleeding_gums_around_natural_teeth"
  "denture_related_sores"
  "dirty_teeth_or_dentures"
  "bad_breath"
  "periodontal_diseases"
  "edentulism"
  "poor_general_oral_health"
  "avoided_eating"
  "uncomfortable_eating"
  "no_oral_health_problems"
  )

# prs
for pheno in "${PHENO[@]}"
do
  Rscript $DATA/PRSice.R \
    --prsice PRSice_linux \
    --base $DATA/base_gwas/gwas.${pheno}.assoc \
    --target $DATA/snp_t2r_noneuropean_sm_qc_maf0.01_updated_sex_target \
    --binary-target T \
    --pheno $DATA/target_noneuropean/${pheno}.txt \
    --cov $DATA/covariates_oral_noneuro.txt \
    --stat OR \
    --or \
    --out $DATA/outcome_noneuro/prs_${pheno}
done
      
```

## Combine prs summary files (base=clsa euro, target=clsa noneuro)

```{r}
  
  oral_prs_target_noneuro <- data.frame()

for (phenotype in phenotypes) {
    file_path <- paste0("~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/outcome_noneuro/prs_", phenotype, ".summary")
    data <- read.table(file_path, header = TRUE)  # assuming the file has headers
    data$Phenotype <- phenotype  # adding a Phenotype column
    oral_prs_target_noneuro <- rbind(oral_prs_target_noneuro, data)
}
write.csv(oral_prs_target_noneuro, "~/Mojan_data/Genetics/Genomics3_clsa/output/oral_health/prs/oral_prs_target_noneuro.csv")

```
